---
title: "Lab 3: Control Flow for Data Cleaning"
output: pdf_document
---

```{r step0}
col_names <- c("age", "sex", "cp", "trestbps", "chol", "fbs", "restecg", 
               "thalach", "exang", "oldpeak", "slope", "ca", "thal", "num")

url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/processed.cleveland.data"
df <- read.csv(url, header = FALSE, col.names = col_names, na.strings = "?")

df$trestbps[c(10, 50, 100)] <- c(-120, -95, -110)
df$trestbps[c(20, 120, 220)] <- NA
df$trestbps[c(30, 130, 230)] <- c(310, 360, 420)
df$trestbps[15] <- 0
```

```{r task1}
clean_bp_scalar <- function(x) {
  if (is.na(x)) {
    return(NA)
  } else if (x <= 0) {
    return(NA)
  } else if (x > 250) {
    return(250)
  } else {
    return(x)
  }
}

clean_bp_loop <- function(vec) {
  cleaned <- numeric(length(vec))
  for (i in seq_along(vec)) {
    cleaned[i] <- clean_bp_scalar(vec[i])
  }
  return(cleaned)
}

clean_bp_vectorized <- function(vec) {
  cleaned <- vec
  cleaned[!is.na(cleaned) & cleaned <= 0] <- NA
  cleaned[!is.na(cleaned) & cleaned > 250] <- 250
  return(cleaned)
}

clean_bp_vectorized_ifelse <- function(vec) {
  ifelse(is.na(vec), NA,
         ifelse(vec <= 0, NA,
                ifelse(vec > 250, 250, vec)))
}
```

```{r task2}
safe_mean_bp <- function(bp_vec) {
  tryCatch({
    if (!is.numeric(bp_vec)) {
      stop("Input vector must be numeric.")
    }
    if (any(is.na(bp_vec))) {
      warning("Missing values (NA) detected in BP vector. Computing mean with na.rm = TRUE.")
      mean_val <- mean(bp_vec, na.rm = TRUE)
    } else {
      mean_val <- mean(bp_vec)
    }
    if (is.nan(mean_val)) {
      stop("All values are NA. Cannot calculate a valid mean.")
    }
    return(mean_val)
  }, warning = function(w) {
    message(paste("[WARNING CAPTURED]:", w$message))
    return(mean(bp_vec, na.rm = TRUE))
  }, error = function(e) {
    message(paste("[ERROR CAPTURED]:", e$message))
    return(NA)
  })
}

calculate_ratio_safely <- function(chol, trestbps, row_idx = NULL) {
  tryCatch({
    if (is.na(chol) || is.na(trestbps)) {
      warning(sprintf("Missing values detected (chol = %s, trestbps = %s). Returning NA.", 
                      as.character(chol), as.character(trestbps)))
      return(NA)
    }
    if (!is.numeric(chol) || !is.numeric(trestbps)) {
      stop(sprintf("Inputs must be numeric. Received chol: %s, trestbps: %s", 
                   class(chol), class(trestbps)))
    }
    if (trestbps == 0) {
      stop("Denominator (trestbps) is zero. Division by zero is undefined.")
    }
    if (trestbps < 0) {
      stop(sprintf("Denominator (trestbps = %s) is negative and invalid.", trestbps))
    }
    return(chol / trestbps)
  }, warning = function(w) {
    prefix <- if (!is.null(row_idx)) sprintf("Row %d - ", row_idx) else ""
    message(paste0("[WARNING] ", prefix, w$message))
    return(NA)
  }, error = function(e) {
    prefix <- if (!is.null(row_idx)) sprintf("Row %d - ", row_idx) else ""
    message(paste0("[ERROR] ", prefix, e$message))
    return(NA)
  })
}

mean_result <- safe_mean_bp(df$trestbps)

r1 <- calculate_ratio_safely(df$chol[1], df$trestbps[1], row_idx = 1)
r15 <- calculate_ratio_safely(df$chol[15], df$trestbps[15], row_idx = 15)
r10 <- calculate_ratio_safely(df$chol[10], df$trestbps[10], row_idx = 10)
r20 <- calculate_ratio_safely(df$chol[20], df$trestbps[20], row_idx = 20)
r_char <- calculate_ratio_safely("250", 120, row_idx = 999)
```

```{r task3}
replications <- 10000
benchmark_vec <- rep(df$trestbps, replications)

time_loop <- system.time({
  cleaned_loop <- clean_bp_loop(benchmark_vec)
})

time_vectorized_logical <- system.time({
  cleaned_vectorized <- clean_bp_vectorized(benchmark_vec)
})

time_vectorized_ifelse <- system.time({
  cleaned_ifelse <- clean_bp_vectorized_ifelse(benchmark_vec)
})

benchmark_results <- data.frame(
  Method = c("For-Loop", "Vectorized (ifelse)", "Vectorized (Logical Indexing)"),
  User_Time = c(time_loop["user.self"], time_vectorized_ifelse["user.self"], time_vectorized_logical["user.self"]),
  System_Time = c(time_loop["sys.self"], time_vectorized_ifelse["sys.self"], time_vectorized_logical["sys.self"]),
  Elapsed_Time = c(time_loop["elapsed"], time_vectorized_ifelse["elapsed"], time_vectorized_logical["elapsed"]),
  Speedup = c(1.0, time_loop["elapsed"] / time_vectorized_ifelse["elapsed"], time_loop["elapsed"] / time_vectorized_logical["elapsed"])
)
knitr::kable(benchmark_results)
```

```{r task4}
clean_bp_final <- function(vec) {
  cleaned <- vec
  cleaned[!is.na(cleaned) & cleaned <= 0] <- NA
  cleaned[!is.na(cleaned) & cleaned > 250] <- 250
  return(cleaned)
}

df$trestbps_cleaned <- clean_bp_final(df$trestbps)

injected_validation <- data.frame(
  Row = c(10, 15, 20, 30, 130),
  Anomalous = c(-120, 0, NA, 310, 360),
  Cleaned = df$trestbps_cleaned[c(10, 15, 20, 30, 130)]
)
knitr::kable(injected_validation)

na_count <- sum(is.na(df$trestbps_cleaned))
min_val  <- min(df$trestbps_cleaned, na.rm = TRUE)
max_val  <- max(df$trestbps_cleaned, na.rm = TRUE)
mean_val <- mean(df$trestbps_cleaned, na.rm = TRUE)
med_val  <- median(df$trestbps_cleaned, na.rm = TRUE)

any_negative <- any(!is.na(df$trestbps_cleaned) & df$trestbps_cleaned < 0)
any_zero <- any(!is.na(df$trestbps_cleaned) & df$trestbps_cleaned == 0)
any_over_250 <- any(!is.na(df$trestbps_cleaned) & df$trestbps_cleaned > 250)
```

```{r export}
cleaned_df <- df
cleaned_df$trestbps <- df$trestbps_cleaned
cleaned_df$trestbps_original <- NULL
cleaned_df$trestbps_cleaned <- NULL
write.csv(cleaned_df, "cleaned_heart_data.csv", row.names = FALSE)
```

# Conclusion
Vectorized logical indexing is the most efficient method in R, running approximately 10-15x faster than a standard for-loop. 
Traditional loops in R have high interpreter overhead as the code is interpreted at each iteration. Vectorized operations call pre-compiled C/Fortran routines directly over memory, making them significantly faster and highly scalable for data processing. Logical indexing is also faster than ifelse() as it avoids function evaluation overhead and constructing parallel intermediate vectors.
