---
title: "Lab 4: Advanced Missing Data Handling"
output: pdf_document
---

# Concept Distinction: NA, NULL, NaN, and "" in R
1. **NA (Not Available)**: Represents a missing value of a specific type (logical, numeric, character) inside a vector or data frame cell.
2. **NaN (Not a Number)**: Represents an undefined mathematical result (e.g., 0/0). It behaves like NA in is.na() checks but is numeric and signifies invalid numeric calculations.
3. **NULL**: Represents the absence of an object or an empty value of length 0. It cannot reside inside data frame cells because columns are vectors of equal length. Assigning NULL to a data frame column deletes the column.
4. **"" (Blank String)**: A valid character vector of length 1 containing zero characters. It is not recognized as missing by R's interpreter (is.na() returns FALSE), though it represents missingness in business logic.

```{r step0}
library(naniar)
library(skimr)
library(ggplot2)

col_names <- c("age", "workclass", "fnlwgt", "education", "education_num", 
               "marital_status", "occupation", "relationship", "race", "sex", 
               "capital_gain", "capital_loss", "hours_per_week", "native_country", "income")

url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data"
df <- read.csv(url, header = FALSE, col.names = col_names, na.strings = "?", strip.white = TRUE)
df <- df[!is.na(df$age), ]

df$age[c(10, 100, 1000)] <- 999
df$age[c(20, 200, 2000)] <- NA
df$workclass[c(30, 300, 3000)] <- ""
df$capital_gain[c(40, 400, 4000)] <- NaN
```

```{r task1}
is_na_demo <- is.na(df$age[20])
is_nan_demo <- is.nan(df$capital_gain[40])
is_null_demo <- is.null(NULL)

test_list <- list(a = 1, b = NULL)
list_null_check <- is.null(test_list$b)

df_null_test <- data.frame(x = 1:5)
df_null_test$x <- NULL
df_columns_after_null <- length(df_null_test)

is_blank_demo <- df$workclass[30] == ""
is_impossible_demo <- df$age[10] == 999

cat("is.na() check (Row 20 Age):", is_na_demo, "\n")
cat("is.nan() check (Row 40 Capital Gain):", is_nan_demo, "\n")
cat("is.null() check (NULL object):", is_null_demo, "\n")
cat("is.null() check (List element):", list_null_check, "\n")
cat("Columns remaining after setting to NULL:", df_columns_after_null, "\n")
cat("Blank string check (Row 30 Workclass):", is_blank_demo, "\n")
cat("Impossible value check (Row 10 Age):", is_impossible_demo, "\n")

naniar::miss_var_summary(df)
```

```{r task2}
df$age[df$age == 999] <- NA

for (col in colnames(df)) {
  if (is.character(df[[col]])) {
    df[[col]][df[[col]] == ""] <- "Unknown"
    df[[col]][is.na(df[[col]])] <- "Unknown"
  }
}

df <- df[!is.nan(df$capital_gain), ]

complete_cases_count <- sum(complete.cases(df))
incomplete_cases_count <- sum(!complete.cases(df))
cat("Complete cases:", complete_cases_count, "\n")
cat("Incomplete cases:", incomplete_cases_count, "\n")
```

```{r task3}
impute_median <- function(x) {
  if (!is.numeric(x)) {
    stop("Input must be a numeric vector.")
  }
  missing_mask <- is.na(x)
  if (any(missing_mask)) {
    med_val <- median(x, na.rm = TRUE)
    x[missing_mask] <- med_val
  }
  return(x)
}

df$age <- impute_median(df$age)
```

```{r task4}
naniar::miss_var_summary(df)
```

```{r task5}
skimr::skim(df)

any_age_999 <- any(df$age == 999, na.rm = TRUE)
any_age_na <- any(is.na(df$age))
any_workclass_blank <- any(df$workclass == "", na.rm = TRUE)

cat("Any age values equal to 999 remaining?", any_age_999, "\n")
cat("Any missing (NA) values in age column?", any_age_na, "\n")
cat("Any blank strings in workclass column?", any_workclass_blank, "\n")

write.csv(df, "cleaned_adult_data.csv", row.names = FALSE)
```

# Interpretation of Cleaning Results
1. **Systematic Identification**: Injected and raw missing data were successfully mapped. Raw missing data (`?` values) were highest in occupation (1,843) and workclass (1,835).
2. **Imputation and Substitution**: Injected impossible age values (999) and NAs in the age variable were resolved via custom median imputation (replaced with median age 37). String missing values and empty cells were converted to "Unknown".
3. **Filtering**: Unrecoverable NaN values in capital_gain (3 observations) were completely removed from the dataset.
4. **Final Validation**: Validation using skimr and direct boolean checks verified that 100% of missing cells in selected columns, blank string entries, and impossible age readings have been successfully treated.
